Databases are an effective and efficient method for storage and access of mass-spectrometry data

### 2025-06-10 manuscript/manuscript.Rmd
Adding edits from Sam

Our previous work showed how the ragged arrays of MS data can be converted into a tidy database table in memory [@Kumler2022] and we now logically extend that method into proper database storage. Here, we test the hypothesis that a “vanilla” implementation of a relational database which exposes the raw m/z and intensity pairs is an intuitive and performant way of storing MS data for exploratory analysis, visualization, and quality control. We compare the time and space required to extract a representative data subset under six conditions and perform these tests on multiple databases as well as mzML and other MS data formats.

@@

Despite the relative simplicity and relevance of our queries, none of the available mass spectrometry (MS) formats had existing functions or examples of all six queries. Unsurprisingly, the mzML file type had the most extensive coverage but documentation and prebuilt functionality was still disappointingly sparse. The pyteomics package provides four "combined examples" that focus on the spectrum visualization and annotation common to proteomics research but provide minimal guidance about chromatogram or retention time range extraction. pyteomics also provides native support for the mzMLb file type and was the only one of the three Python packages to do so, deserving praise for the minimal disruption that mzMLb files placed on existing pipelines if they were to switch from mzML to mzMLb. The pyopenms package provides similarly extensive documentation for proteomic and scan-based analysis but again lacks information about subsetting in the retention time direction, though the existence of an undocumented parser (`get2DPeakDataLong`) provides a simple way to do this for MS<sup>1</sup> data. Additionally, pyOpenMS required installing an old version of the package (3.0.0), Python itself (3.11) and the numpy package (<2.0) due to more recent builds requiring AVX support which was unavailable on our hardware. pymzml is intentionally a lightweight parser focused exclusively on reading mzML files but does not supply any functions for the queries other than scan extraction by number and the "Spectrum and Chromatogram" documentation module was empty at the time of writing (February 2025).

@@

mzDB access was hamstrung by several issues, primarily the outdated repository that implies Python and R support via a port from Rust but was unavailable at the time of development, though we are grateful for the responsive developer who notified us that this implementation was not feature-complete. This required that we deduce the SQLite BLOB type compression format from scratch when writing a parser and spend extensive time reading through the documentation to determine how best to link the various tables provided in the mzDB file. Scan metadata in this file type is stored as raw XML strings, producing the worst of both worlds in requiring both SQLite knowledge in their extraction and XML processing to obtain the relevant information. Additionally, its failure to implement the clever bounding box and run slice scheme for MS/MS data negated with our ability to avoid parsing every MS/MS spectrum when performing precursor and fragment searches.

@@

The simplest and most abundantly documented query was the extraction of a single spectrum. In many ways, this is the fundamental unit of mass spectrometry and thus many formats are highly optimized for its extraction into manipulatable data (Figure 2A and 2D). Here, we found that the mzML and mzMLb file types were consistently the slowest to parse and required multiple seconds, likely highlighting inefficiencies in the `pyteomics` package used to parse both file types. `pyopenms` also struggled to open and extract a specific scan, requiring several seconds due in large part to the expensive initiation function, after which requests were orders of magnitude faster (Supplemental Figure XX). It is also worth noting that while both of these packages provided rapid extraction of a *random* spectrum, a significant overhead was introduced by needing to scan through the file to find a *specific* spectrum by scan number as scans are not always consecutive (e.g. during multiple MS experiments, polarity-switching runs, or when scans have been filtered out during conversion) and no metadata was obviously available that would have allowed using the index directly to a specific scan.

@@

Most methods were linear extrapolations of the single chromatogram numbers shown above as expected from a simple for loop, with notable exceptions for pyteomics (and thus the mzMLb format), the 2D peak method of pyopenms, and DuckDB (Figure 3). pyteomics and pyopenms both had significant overhead upon initial load that resulted in faster query times afterward that performed better than a 1:1 extrapolation would expect, with pyopenms matching SQLite's speed after 10 chromatograms and Parquet's speed after 30. Interestingly, the methods that had a best-fit linear slope less than 1:1 also all had exponential fits, with performance at high chromatogram number actually worse than would be expected from a prediction fit to the timings for 1 and 3 chromatogram extractions (Supplemental figure XX).

@@

Finally, a complexity penalty must be noted for formats and packages requiring complex installation procedures. pyopenms appears to be the worst offender here, with its bindings to the OpenMS C++ libraries requiring us to step back to Python 3.11 and numpy 1.26 to successfully access the data on our setup and required direct input from the maintainers. The Java applications for MzTree and, more egregiously mzMD, provided essentially zero documentation on installation consisting of a single README file without an intact HTTP request example and have not seen updates in years. Similarly, mzDB files were difficult to parse due to its opaque SQLite schema and use of the BLOB encoding type that again lacked examples or documentation outside of Java and had to be deduced iteratively. Of course, these methods eventually provided enough information that they could be parsed unlike Aird, toffee, and UIMF.

@@

Wilhelm et al. (2012) @Wilhelm2012 performed comparisons between mz5 and mzML with results indicating that mz5 was three times faster than mzML parsers. Their values of 0.16 seconds per million *m/z*/intensity arrays correspond to an estimated query time of 0.13 milliseconds which is faster than our measured 3 milliseconds, though our mz5 file was approximately four times smaller (20% mzML size) instead of two times smaller (50%).

@@

The mzMLb group [@Bhamber2021] reports only info for spectrum access at approximately 15ms/scan which agrees with our 5-100ms/scan estimates only if the data is loaded ahead of time. They perform extensive comparisons to mzML at varying compression methods and levels but our default options of 1024 KB chunk sizes for the mzMLb file and zlib but not numpress compression for the mzML resulting in timing values very similar between mzML and mzMLb using the pyteomics library for access. They also compared to the mz5 file type and we are able to validate their results of the mzML+zlib occupying significantly more space, though our mz5 parser outperformed theirs for full scan and MS/MS data by two orders of magnitude in time.

We were also unable to test several other recent and promising formats. Aird does not report full query times for any of the metrics reviewed here, though they claim their StackZDPD algorithm [@Wang2022] can improve decompression speed by three times and that the file size is 54% of the vendor file.[@Lu2022] Similarly, the toffee format for time-of-flight DIA data reports sizes about equal to vendor or 60% of centroid mzml + numpress with query speeds 4 times faster for scans (spectrum-centric, 2 seconds for mzML, 0.5 for toffee) and 100 times faster for chroms (peptide-centric, 168 seconds for mzML, 1.8 for toffee).[@Tully2020] The Unified Ion Mobility Format (UIMF, @Beagley2009) from Pacific Northwest National Laboratory format did not report direct comparisons to any of the available formats and thus we must remain unclear on its performance capabilities.

>>>>

Our previous work showed how the ragged arrays of MS data can be converted into a tidy database table in memory [@Kumler2022] and we now logically extend that method into proper database storage. Here, we test the hypothesis that a “vanilla” implementation of a relational database which exposes the raw m/z and intensity pairs is an intuitive and performant way of storing MS data for exploratory analysis, visualization, and quality control. We compare the time and space required to extract a representative data subset under six conditions and perform these tests on multiple databases as well as mzML and other MS data formats. Our specific questions were:

1. Is there a simple database schema that enables exploratory MS queries with basic SQL statements?
2. How expensive in time and (disk) space is it to access MS data in SQLite, DuckDB, and Parquet formats?
3. How does this cost compare to more complex formats that have been previously proposed?

@@

Despite the relative simplicity and relevance of our queries, none of the available mass spectrometry (MS) formats had existing functions or examples of all six queries. Unsurprisingly, the mzML file type had the most extensive coverage but documentation and prebuilt functionality was still disappointingly sparse. The pyteomics package provides four "combined examples" that focus on the spectrum visualization and annotation common to proteomics research but provide minimal guidance about chromatogram or retention time range extraction. Pyteomics also provides native support for the mzMLb file type and was the only one of the three Python packages to do so, deserving praise for the minimal disruption that mzMLb files placed on existing pipelines if they were to switch from mzML to mzMLb. The pyopenms package provides similarly extensive documentation for proteomic and scan-based analysis but again lacks information about subsetting in the retention time direction, though the existence of an undocumented parser (`get2DPeakDataLong`) provides a simple way to do this for MS<sup>1</sup> data. Additionally, pyOpenMS required installing an old version of the package (3.0.0), Python itself (3.11) and the numpy package (<2.0) due to more recent builds requiring AVX support which was unavailable on our hardware. Pymzml is intentionally a lightweight parser focused exclusively on reading mzML files but does not supply any functions for the queries other than scan extraction by number and the "Spectrum and Chromatogram" documentation module was empty at the time of writing (February 2025).

@@

mzDB access was hamstrung by several issues, primarily the outdated repository that implies Python and R support via a port from Rust but was unavailable at the time of development, though we are grateful for the responsive developer who notified us that this implementation was not feature-complete. This required that we deduce the SQLite BLOB type compression format from scratch when writing a parser and spend extensive time reading through the documentation to determine how best to link the various tables provided in the mzDB file. Scan metadata in this file type is stored as raw XML strings, producing the worst of both worlds in requiring both SQLite knowledge in their extraction and XML processing to obtain the necessary information. Additionally, mzDB seems to dump all MS/MS data into a single bounding box, meaning that we were unable to use the scheme to avoid parsing every MS/MS spectrum when performing precursor and fragment searches.

@@

The simplest and most abundantly documented query was the extraction of a single spectrum. In many ways, this is the fundamental unit of mass spectrometry and thus many formats are highly optimized for its extraction into manipulatable data (Figure 2A and 2D). Here, we found that the mzML and mzMLb file types were consistently the slowest to parse and required multiple seconds, likely highlighting inefficiencies in the `pyteomics` package used to parse both file types. `pyopenms` also struggled to open and extract a specific scan, requiring several seconds due in large part to the expensive initiation function, after which requests were orders of magnitude faster (Supplemental Figure XX). It is also worth noting that while both of these packages provided rapid extraction of a *random* spectrum, a significant overhead was introduced by needing to scan through the file to find a *specific* spectrum by scan number. Scans are not always consecutive (e.g. during multiple MS experiments, polarity-switching runs, or when scans have been filtered out during conversion) and no metadata was obviously available that would have allowed using the index directly to a specific scan number.

@@

Most methods were linear extrapolations of the single chromatogram numbers shown above as expected from a simple for loop, with notable exceptions for pyteomics (and thus the mzMLb format), the 2D peak method of pyopenms, and DuckDB (Figure 3). Pyteomics and pyopenms both had significant overhead upon initial load that resulted in faster query times afterward that performed better than a 1:1 extrapolation would expect, with pyopenms matching SQLite's speed after 10 chromatograms and Parquet's speed after 30. Interestingly, the methods that had a best-fit linear slope less than 1:1 also all had exponential fits, with performance at high chromatogram number actually worse than would be expected from a prediction fit to the timings for 1 and 3 chromatogram extractions (Supplemental figure XX).

@@

Finally, a complexity penalty must be noted for formats and packages requiring complex installation procedures. Pyopenms appears to be the worst offender here, with its bindings to the OpenMS C++ libraries requiring us to step back to Python 3.11 and numpy 1.26 to successfully access the data on our setup and required direct input from the maintainers. The Java applications for MzTree and, more egregiously mzMD, provided essentially zero documentation on installation consisting of a single README file without an intact HTTP request example and have not seen updates in years. Similarly, mzDB files were difficult to parse due to its opaque SQLite schema and use of the BLOB encoding type that again lacked examples or documentation outside of Java and had to be deduced iteratively. Of course, these methods eventually provided enough information that they could be parsed unlike Aird, toffee, and UIMF.

@@

Wilhelm et al. (2012) @Wilhelm2012 performed comparisons between mz5 and mzML with results indicating that mz5 was three times faster than mzML parsers. Their values of 0.16 seconds per million *m/z*/intensity arrays correspond to an estimated query time of 0.13 milliseconds which is faster than our measured 3 milliseconds, though our mz5 file was approximately four times smaller (20% mzML size) instead of half the mzML size.

@@

The mzMLb group [@Bhamber2021] reports only info for spectrum access at approximately 15ms/scan which agrees with our 5-100ms/scan estimates only if the data is loaded ahead of time. They perform extensive comparisons to mzML at varying compression methods and levels but we stuck with the default options of 1024 KB chunk sizes for the mzMLb file and zlib-only compression for the mzML. This resulted in timing values very similar to those they reported when using the pyteomics library for access. They also compared to the mz5 file type and we are able to validate their results of the mzML+zlib occupying significantly more space, though our mz5 parser outperformed theirs for full scan and MS/MS data by two orders of magnitude in time.

We were also unable to test several other recent and promising formats. Aird does not report full query times for any of the metrics reviewed here, though they claim their StackZDPD algorithm [@Wang2022] can improve decompression speed by three times and that the file size is 54% of the vendor file.[@Lu2022] Similarly, the toffee format for time-of-flight DIA data reports sizes about equal to vendor or 60% of centroid mzml + numpress with query speeds 4 times faster for scans (spectrum-centric, 2 seconds for mzML, 0.5 for toffee) and 100 times faster for chroms (peptide-centric, 168 seconds for mzML, 1.8 for toffee).[@Tully2020] The Unified Ion Mobility Format (UIMF) @Beagley2009 from Pacific Northwest National Laboratory format did not report direct comparisons to any of the available formats and thus we must remain unclear on its performance capabilities.
